Optical characterization of gaps in directly bonded Si compound optics using infrared spectroscopy

### 2015-01-06 AO_bondedSi.tex
DTJ edits up to page 18

\subsection{Predicted transmission spectrum for bonded Si with finite interface gap}
\label{secTheory}
There is a key hurdle in computing the transmission spectrum for bonded Si with a finite interface gap- multiple incoherent reflections occur within the Si substrate at normal incidence.  Treatments for multiple coherent reflections have been described in detail in the optics literature \verb|\cite{2007fuph.book.....S}|.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  We adapted the wave transfer technique for incoherent interactions \verb|\cite{2002ApOpt..41.3978K}|.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si sample with no gap, and second a pair of bonded Si samples with a gap thickness $d$.  The DSP Si sample with no air gap has a transmission equal to:

@@

which has an average value of about 53\%.  Note that this transmission is above a na\"ive value of $T_{DSP}=(1-R^2)$, which does not take into account multiple incoherent reflections, and is therefore an underestimate.  Figure \ref{figAbsoluteTrans} shows a plot of Equation \ref{eqnAbsDSPtrans} in the wavelength range 1200$-2500$ nm.  If the Si wafer path thickness is less than the coherence length for the given spectral bandwidth, then the multiple reflections would interfere coherently, and the Si wafer would behave as a Fabry-P\`erot etalon.  This scenario is beyond the scope of this paper, since we are primarily interested in thick substrates with small gaps in between the interfaces.%We work out this behavior in the Appendix and show that in the limit $l>> \lambda^2 / \delta \lambda $ the spectral transmission tends to equation X.

@@

It is clear that in the limit $\lim_{d \rightarrow 0}$, the etalon approaches 100\% transmission.  Figure \ref{figEtalonRelTrans} shows a plot of Equation \ref{eqFP} for gap sizes, $d$, of 0, 20, 35, 60, 350, and 3500 nm.  In Section \ref{secResults} we compare wavelength dependent transmission measurements to by-eye fits of Equation \ref{eqFP}.

\section{Imbedding gaps of known sizes}

@@

The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify substrates with thicknesses larger than 1 mm as ``pucks''.  Substrates with thickness less than 1.0 mm thick are called ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \verb|\cite{2007ApOpt..46.6793H}| than in wafers.  Our ultimate goal is to bond pucks with thicknesses as large as 30 mm thick.  These pucks truly resemble over-sized hockey pucks.

The three important characteristics of the gaps are their axial extent, their lateral areal extent, and their areal fill factor.  We bored holes in silicon with inductively coupled plasma etching.  We used two different gases with low and high etch rates.  CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  The area exposed to the process gases was about 1 inch diameter circles, or 15 mm $\times$ 20 mm rectangles.  We protected the unexposed areas with thick photoresist.  We achieved etch depths of 15$\pm$5 nm and 4000$\pm$100 nm, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.  Table \ref{tabbondexper} has more information about the bonding samples, and Table \ref{tabPatternFills} lists the properties of the patterns.  The fill factor is defined as the pattern area covered by gaps relative to the total pattern area.  In a second round of experiments we achieved gap sizes of 14$-$95 nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profiliometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.  Table \ref{tabbondexper} lists part numbers and descriptions for our bonded substrates.

@@

\section{Substrate preparation}
We cleaned the surfaces before bonding to minimize the interfacial particle density.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer has a 2 inch diameter beam at $\lambda=$632.8 nm.  We found a typical peak to valley surface flatness of 2 waves over the central 2 inch diameter.  We prepared the substrates with standard cleaning procedures of solvents in a megasonic.  We then applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with an N$_2$ gun.  We pressed the Si substrates together from the center to the outside.

\section{Measurement strategy, conditions, and details}
We took transmission spectra of bonded Si wafers and pucks, and double side polished Si reference wafers. The spectrophotometer is a Agilent Cary 5000 UV-Vis-NIR.  The typical measurement settings and tool performance are listed in Table \ref{tab:Cary5000pars}.  More details about the tool are available at the vendor's website.  Briefly, the tool has a double monochromator.  The double monochromator feature reduces scattered light.  The vendor reported linearity exceeds 40 dB.

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  We experimented with single front and double modes.  Double mode takes a reference spectrum simultaneously to the target spectrum.  In both single and double beam mode, a baseline scan is taken with no sample in the measurement holder.  The baseline spectrum is the raw spectral response of the system.  We automatically divide sample spectra by the baseline spectrum.  In single beam mode, the baseline measurements must be repeated about every half hour.  We estimated the uncertainty and measurement repeatability of single and double modes in the following way.  We computed the RMS error and measurement repeatability in a single measurement by computing the standard deviation of a spectrum taken immediately after a baseline scan with no sample present.  The predicted mean level should naturally be 100\%.  We found that the mean value in single mode drifted by about 0.2\% over several measurements.  The standard deviation of the featureless spectrum was typically 0.02\% from point to point, with some portions of the spectrum demonstrating increased variance, perhaps attributable to atmospheric absorption; we list the wavelengths of instrumental artifacts and above-average variance in Table \ref{tab:Cary5000pars}. In double beam mode, the spectrograph performance was much better.  We put a double side polished (DSP) Si wafer in the reference beam, and no sample in the target beam during the baseline scan.  In this way the transmission spectrum of a target Si wafer is automatically normalized.  The double beam mode effectively divides out the 0.2\% drift in the mean transmission level attributable to lamp drift, as seen in single beam measurements.  In double beam mode, the drift in mean transmission level is merely 0.02\%.  These values are accurate for measurements over a half hour timescale or less.

The choice of spectral sampling was straightforward.  The spectral sampling was 2.0 nm.  The spectral resolution was 5.0 nm, the lowest spectral resolution the Cary5000 offers.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\`erot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  Such large gaps would be easily detectable by other means, like IR imaging.

@@

We looked for the indication of large gaps detectable as ``IR bubbles'' \verb|\cite{1992JEMat..21..669M}| with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the detector artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si samples on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  Naturally, the unbonded DSP wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrates as discussed in the literature (see Section \ref{secHistory}).  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

@@

\subsection{Constructing the predicted transmission}
We constructed the predicted transmission in the following way.  If IR imaging was unavailable, our optimistic prior belief was simply that the bond should be perfect and indistinguishable from a solitary DSP Si wafer, (i.e. 100\% relative transmission).  For the cases where we had bored gaps of known dimension and fill factors we could predict the transmission with the Fabry-Perot machinery described in section \ref{secTheory}.  Specifically we apply Equation \ref{eqFP}.  For non-unit fill factor, we simply average 100\% transmission with the predicted gap transmission, weighted by the estimated fill factor listed in Table \ref{tabPatternFills}.  No attempt was made to verify the delivered fill factor, but since high precision lithographic processes were employed, we assume the delivered fill factor is equal to the designed fill factor with no uncertainty.  The depth of the meshes has an associated uncertainty listed in Table \ref{tabbondexper}.  We incorporated the uncertainty in the mesh depth in the following way.  We computed the transmission curve for the 1$\sigma$ lower and upper bounds, and plot a filled band marked as the prediction band.  The uncertainties in the mesh depths were low enough that the band spread includes all intermediate lines of predicted transmission between plus and minus 1$\sigma$.  In cases where the IR imaging was unavailable the mesh position was only coarsely known, so our prediction band had to incorporate our uncertainty about whether or not our measurement beam was on or off or overlapping the mesh border.  If the beam was off the mesh we assumed 100\% relative transmission, so we simply set the lower bound of the predicted transmission to the equivalent of a zero gap.

\subsection{Spectra of direct bonded Si with known gap sizes}
Figures \ref{VS21specm1} to \ref{VS21specf1} show examples of the predicted and measured IR spectra.  Figure \ref{VS21specm1} shows an example spectrum in which the measurement is consistent with the prediction for a gap with the known conditions.  Meanwhile, Figure \ref{VS21specf1} shows a spectrum in which the measurement marginally exceeds the prediction (by about $\sim 2 \sigma$).  Other parts show less consistency with the prediction.  A complete description of the diversity of the measured spectra requires an appeal to optical phenomena beyond the scope of this paper.  Briefly we classify the diversity of measurements into the categories laid out in the table below.

@@

\subsection{Mitigation of measurement artifacts}
We mitigated measurement artifacts in the following ways.  For the Sharpie marker behaving as an anti-reflection (AR) coating, we simply removed the sharpie with IPA.  We have mostly been able to rule out wavelength-dependent diffraction from the mesh pattern as a measurement artifact, since the beam size is large and the angular spread of diffraction from the mesh pattern would be much smaller than the spread of the beam.  Still, to remove this concern, we moved to using the coarse mesh rather than the medium or fine meshes, when possible.  We saw no difference in the bond quality surrounding the medium or fine meshes, which is consistent with results from the literature \verb|\cite{1992JEMat..21..669M}|.  Specifically, the gas pressure of bonded wafers grows as the fill factor of gaps decreased.  In other words, what matters is the fill factor of gaps, not their relative sizes.  Since the fine, medium, and coarse meshes had the same fill factors, we anticipate no difference in the pressure of the gaps, and therefore no difference in the bonding properties.

>>>>

\subsection{Predicted transmission spectrum for bonded Si with finite interface gap}
\label{secTheory}
A pair of bonded silicon wafers or pucks with a small gap actually forms there coupled cavities with the air gap as the the central cavity.  The outer cavities, those within the Si, however, were not produced with coherence in mind.  Their sides are not necessarily flat and parallel to the required level and the specifics of their deviation from perfect etalons are usually not known to the level necessary to treat them as oherent structures in the analysis.  The multiple internal reflections within the Si therefore present a hurdle to an analysis of the air gaps where the gap size si small enough that the cavity it forms can always be treated as a coherent structure.  Treatments for multiple coherent reflections have been described in detail in the optics literature \verb|\cite{2007fuph.book.....S}|.  The wave transfer matrix technique treats each dielectric interface as a matrix with elements relating the pre- and post- interface complex amplitudes in the left and right directions.  We adapted the wave transfer technique for incoherent interactions \verb|\cite{2002ApOpt..41.3978K}| to deal with the cavities within the Si.  Specifically I constructed an incoherent wave transfer matrix whose elements relate the intensities (and not complex amplitudes) before and after an interface.  The details are worked out in Appendix \ref{sec:Append-IMRTMM}.  We computed the transmission for two scenarios- first a double side polished (DSP) Si sample with no gap, and second a pair of bonded Si samples with a gap thickness $d$.  The DSP Si sample with no air gap has a transmission equal to:

@@

which has an average value of about 53\%.  Note that this transmission is above a na\"ive value of $T_{DSP}=(1-R^2)$, which does not take into account multiple incoherent reflections, and is therefore an underestimate.  Figure \ref{figAbsoluteTrans} shows a plot of Equation \ref{eqnAbsDSPtrans} in the wavelength range 1200$-2500$ nm.  If the Si wafer path thickness is less than the coherence length for the given spectral bandwidth, then our assumptions break down and the multiple reflections would interfere coherently, and the Si wafer would behave as a Fabry-P\`erot etalon.  This scenario is beyond the scope of this paper, since we are primarily interested in thick substrates with small gaps in between the interfaces.

@@

It is clear that in the limit $\lim_{d \rightarrow 0}$, the etalon approaches 100\% transmission.  Figure \ref{figEtalonRelTrans} shows a plot of Equation \ref{eqFP} for gap sizes, $d$, of 0, 20, 35, 60, 350, and 3500 nm.  In Section \ref{secResults} we compare wavelength dependent transmission measurements to by-eye fits of Equation \ref{eqFP}.  Depending upon the intended operating wavelength, gaps as small as 35-60 nm can cause losses comparable to or greater than the losses produced by antireflection coatings.

\section{Laboratory demonstration: Imbedding gaps of known sizes}

@@

The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify substrates with thicknesses larger than 1 mm as ``pucks''.  Substrates with thickness less than 1.0 mm thick are called ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is slower in thick pucks \verb|\cite{2007ApOpt..46.6793H}| than in wafers.  Our ultimate goal is to bond optics with thicknesses as large as 30 mm.  These pucks truly resemble over-sized hockey pucks.

The three important characteristics of the gaps are their axial extent, their lateral areal extent, and their areal fill factor.  We bored holes in silicon with inductively coupled plasma etching.  We used two different gases with low and high etch rates.  CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  The area exposed to the process gases was either a 25 mm diameter circle, or a 15 mm $\times$ 20 mm rectangle.  We protected the unexposed areas with thick photoresist.  We achieved etch depths of 15$\pm$5 nm and 4000$\pm$100 nm, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.  Table \ref{tabbondexper} has more information about the bonding samples, and Table \ref{tabPatternFills} lists the properties of the patterns.  The fill factor is defined as the pattern area covered by gaps relative to the total pattern area.  In a second round of experiments, we achieved gap sizes of 14$-$95 nm on four substrates with three meshes with coarse, medium, and fine boxes, each with 50\% fill factors.  For measurements taken with the Veeco NT9100 Optical Profiler, uncertainties were constructed by inspecting the histogram of topology, as shown in Figure \ref{figVS20pattern}.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profiliometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.  Table \ref{tabbondexper} lists part numbers and descriptions for our bonded substrates.

@@

We cleaned the surfaces before bonding to minimize the interfacial particle density.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer used for these measurements has a 50 mm diameter beam at $\lambda=$632.8 nm.  We found a typical peak to valley surface flatness of 2 waves over the central 50 mm diameter.  We prepared the substrates with standard cleaning procedures of solvents in a megasonic.  We then applied MHz frequency oxygen plasma ashing.  We soaked the wafers in DI water then dried with N$_2$.  We pressed the Si substrates together from the center to the outside.

\section{Measurement strategy, conditions, and details}
We took transmission spectra of bonded Si wafers and pucks, and double side polished Si reference wafers using an Agilent Cary 5000 UV-Vis-NIR spectrophotometer.  Table \ref{tab:Cary500pars} lists the typical measurement settings and tool performance.  The tool has a double monochromator, which reduces scattered light.  The vendor reported linearity exceeds 40 dB.

The Cary 5000 has four operation modes: single front, single back, double, and double reverse.  We experimented with single front and double modes.  Double mode takes a reference spectrum simultaneously with the target spectrum.  In both single and double mode, a baseline scan is taken with no sample in the measurement holder.  The baseline spectrum is the raw spectral response of the system.  We automatically divide sample spectra by the baseline spectrum.  In single beam mode, the baseline measurements must be repeated about every half hour for post-process division.  In double beam mode the front beam is automatically divided by the rear beam, which corrects out lamp luminosity drift, for example.  We estimated the uncertainty and measurement repeatability of single and double modes in the following way: we computed the RMS error and measurement repeatability in a single measurement by computing the standard deviation of a spectrum taken immediately after a baseline scan with no sample present.  We found that the mean value in single mode drifted by about 0.2\% over several measurements.  The standard deviation of the featureless spectrum was typically 0.02\% from point to point, with some portions of the spectrum demonstrating increased variance, perhaps attributable to atmospheric absorption.  In double beam mode, the spectrograph drift was much lower.  The simultaneous reference in double beam mode divides out the 0.2\% drift in the mean transmission level attributable to lamp drift that we saw in single beam measurements.  In double beam mode, the drift in mean transmission level is 0.02\%.  These values are accurate for measurements over a half hour timescale or less.  With accuracies and drift rates at this level, the curves in Figure \ref{figEtalonRelTrans} indicate that the axial extent of gaps as small as a few nm should be measurable.

The spectral sampling was 2.0 nm.  We used a spectral resolution of 5.0 nm, the lowest spectral resolution the Cary5000 offers.  We expect smoothly varying spectral features for all conceivable interface gaps in Si bonded wafers.  Specifically, Fabry-P\`erot fringes attributable to interfacial gaps in bonded Si will be well-sampled with 5 nm spectral resolution, so long as the interface gap is less than $\sim$100 $\mu$m.  The axial extends of such large gaps would be easily measurable by other means, like IR imaging.

@@

Before testing the bonded pieces from Table 1 using spectrophotometry, we looked for the indication of large gaps detectable as ``IR bubbles'' \verb|\cite{1992JEMat..21..669M}| with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  The combination of the absorption of Si, the lamp spectrum and the detector responsivity led to an effective bandpass from approximately 1.15$-$1.6 $\mu$m.  The field of view was smaller than the 100 mm diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the detector artifacts since the bubbles translate with the sample when we dither its position.  We detected IR bubbles in all 3 of the bonded Si samples on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  As expected, the unbonded DSP wafers show no such IR bubbles.  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

@@

\subsection{Constructing the predicted transmission}
We constructed the predicted transmission  for different artificially induced gaps of known dimension and fill factors using the Fabry-P\`{e}rot machinery described in Section \ref{secTheory}, in particular Equation \ref{eqFP}.  For non-unit fill factor, we assume a perfect bond outside of the artifical gap region and average 100\% transmission with the predicted gap transmission, weighted by the estimated fill factor listed in Table \ref{tabPatternFills}.  No attempt was made to verify the delivered fill factor, but since high precision lithographic processes were employed, we assign no uncertainty to the difference between the delivered fill factor and the designed fill factor.  The depth of the meshes has an associated uncertainty listed in Table \ref{tabbondexper}.  We incorporated the uncertainty in the mesh depth in the following way.  We computed the transmission curve for the 1$\sigma$ lower and upper bounds, and plot a filled band marked as the prediction band.  In some cases the IR imaging was unavailable and we had to place the spectrometer aperture on the mesh without being able to confirm the exact location of the meshed area of the substrate.  Our prediction band had to incorporate our uncertainty about whether or not our measurement beam was overlapping the mesh border.  This overlap behaves like a fill factor weighted towards 100\% bonded.

\subsection{Spectra of direct bonded Si with known gap sizes}
Figures \ref{VS21specm1} to \ref{VS21specf1} show examples of the predicted and measured IR spectra.  Figure \ref{VS21specm1} shows an example spectrum in which the measurement is consistent with the prediction for a gap with the known conditions.  Meanwhile, Figure \ref{VS21specf1} shows a spectrum in which the measurement marginally exceeds the prediction (by about $\sim 2 \sigma$).  Other parts show less consistency with the prediction.

@@

\subsection{Mitigation of measurement artifacts}
We mitigated measurement artifacts in the following ways.  We have mostly been able to rule out wavelength-dependent diffraction from the mesh pattern as a measurement artifact, since the beam size is large and the angular spread of diffraction from the mesh pattern would be much smaller than the spread of the beam.  Still, to remove this concern, we moved to using the coarse mesh rather than the medium or fine meshes, when possible.  We saw no difference in the bond quality surrounding the medium or fine meshes, which is consistent with results from the literature \verb|\cite{1992JEMat..21..669M}|.  Specifically, the gas pressure of bonded wafers grows as the fill factor of gaps decreased.  In other words, what matters is the fill factor of gaps, not their relative sizes.  Since the fine, medium, and coarse meshes had the same fill factors, we anticipate no difference in the pressure of the gaps, and therefore no difference in the bonding properties.
